FLM for growth as function of SPEI

### 2021-01-25 analysis/paper/vital rate models.Rmd
calculate size_prev instead of size_next and switch models to use size_prev as covariate instead of size.  This means SPEI history starts at the month of the census when the response variables were recorded.

## Prep precip data
Get only rows for february (the month of the census), and year for joining

@@

```{r}
unique(xa2$lat)
mean(c(-2.25, -2.5))
```
So, from -2.25 to -2.5

@@

```{r}
demog_full <-
  demog_full %>% 
  filter(ht < 200, ht_next < 200) %>% 
  filter(shts > 0, shts_next > 0, ht > 0, ht_next > 0) %>% 
  mutate(size = shts*ht, size_next = shts_next*ht_next, log_size = log(size), log_size_next = log(size_next))
```

@@

Fit the GAM with crossbasis smooth for lagged drought

```{r}
g_post_cf <-
  bam(log_size_next ~ 
        s(log_size) + 
        s(year_fac, bs = "re") + 
        s(plot, bs = "re") + #random effect of plot
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal-ish response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      # family = gaussian(link = "identity"),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog_post_cf,
      method = "REML",
      cluster = cl)
```

@@

```{r}
summary(g_post_cf)
```
Remember, R-sq is probably so high because size this year is a very good predictor of size next year.

@@

```{r}
g_post_1ha <-
  bam(log_size_next ~ 
        s(log_size) +
        s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog_post_1ha,
      method = "REML",
      cluster = cl)
```

@@

```{r}
summary(g_post_1ha)
```
```{r}
draw(g_post_1ha)
```

@@

```{r}
sr <- range(demog_post_cf$log_size)

newdata_covar_cf <- 
  tibble(log_size = seq(sr[1], sr[2], length.out = 100),
         plot = "newplot",
         year_fac = "newyear", 
         spei_history = mean(demog_post_cf$spei_history),
         L = demog_post_cf$L[1:100, ])

@@

covar_growth_cf <-
  ggplot(out, aes(x = exp(log_size), y = exp(.fitted))) + 
  geom_point(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.1) +
  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  # geom_rug(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.25) +
  theme_linedraw() +
  labs(x = TeX("size_t_"), y = TeX("size_{t+1}_"))
covar_growth_cf
ggsave(here("analysis", "figures", "growth_cf_size.png"), covar_growth_cf)
```
```{r}
sr <- range(demog_post_1ha$log_size)
newdata_covar_1ha <- 
  tibble(log_size = seq(sr[1], sr[2], length.out = 100),
         plot = "newplot",
         year_fac = "newyear",
         spei_history = mean(demog_post_1ha$spei_history),
         L = demog_post_cf$L[1:100, ])

@@

covar_growth_1ha <-
  ggplot(out, aes(x = exp(log_size), y = exp(.fitted))) + 
  geom_point(data = demog_post_1ha, aes(x = size, y = size_next), alpha = 0.1) +
  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  # geom_rug(data = demog_post_1ha, aes(x = size, y = size_next), alpha = 0.25) +
  theme_linedraw() +
  labs(x = TeX("size_t_"), y = TeX("size_{t+1}_"))

covar_growth_1ha
ggsave(here("analysis", "figures", "growth_1ha_size.png"), covar_growth_1ha)
```

@@

```{r}
# use mean size plant across both habitats so plots are comparable
ref_data <-
  demog_post %>%
  dplyr::summarize(log_size = log(mean(size, na.rm = TRUE)),  #log of mean size, not mean of log_size
                   year_fac = "newyear",
                   plot = "newlevel")

mean_size <- exp(ref_data$log_size)
```

@@

```{r}
s_post_cf <-
  bam(surv ~ 
        s(log_size) +
        s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      method = "REML",
      cluster = cl)
```
```{r}
# gam.check(s_post_cf) #not sure what to expect from any of these plots
appraise(s_post_cf)
```
```{r}
summary(s_post_cf)
```

@@

```{r}
s_post_1ha <-
  bam(surv ~ 
        s(log_size) +
        s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      method = "REML",
      cluster = cl)
```
```{r}
summary(s_post_1ha)
```

@@

covar_s_cf <-
  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_cf) +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  labs(x = TeX("size_t_"), y = "P(survival)") +
  theme_linedraw()

@@

covar_s_1ha <-
  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_1ha) +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  labs(x = TeX("size_t_"), y = "P(survival)") +
  theme_linedraw()

@@

Around 0.05 percent of plants flowered.

```{r}
f_post_cf <-
  bam(flwr ~ 
        s(log_size) +
        s(year_fac, bs = "re") + 
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      method = "REML",
      cluster = cl)
```

@@

```{r}
appraise(f_post_cf)
# concurvity(f_post_cf, full = FALSE)
```
these residuals are troubling

@@

```{r}
f_post_1ha <-
  bam(flwr ~ 
        s(log_size) +
        s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      method = "REML",
      cluster = cl)
```

@@

## Covariate plots
Effect of size with all else held average.

@@

covar_f_cf <-
  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
  stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = demog_post_cf) +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  labs(x = TeX("size_t_"), y = "P(flowering)") +
  theme_linedraw()

@@

covar_f_1ha <-
  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
  stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = demog_post_1ha) +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  labs(x = TeX("size_t_"), y = "P(flowering)") +
  theme_linedraw()

covar_f_1ha
ggsave(here("analysis", "figures", "flwr_1ha_size.png"), covar_f_1ha)
```
## Get plot data

>>>>

## Prep precip data

Get only rows for february (the month of the census), and year for joining

@@

```{r}
unique(xa2$lat)
mean(c(-2.25, -2.5))
```

So, from -2.25 to -2.5

@@

```{r}
demog_full <-
  demog_full %>% 
  filter(ht < 200, ht_prev < 200) %>% 
  filter(shts > 0, shts_prev > 0, ht > 0, ht_prev > 0) %>% 
  mutate(size = shts*ht, size_prev = shts_prev * ht_prev, log_size = log(size), log_size_prev = log(size_prev))
```

@@

Fit the GAM with crossbasis smooth for lagged drought.

Preliminary modeling showed that if year is included as random effect it is concurve with SPEI history (which would mean biased p-values and coefficients).

```{r}
g_post_cf <-
  bam(log_size ~ 
        s(log_size_prev) + 
        # s(year_fac, bs = "re") + 
        s(plot, bs = "re") + #random effect of plot
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal-ish response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      # family = gaussian(link = "identity"),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog_post_cf,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

@@

```{r}
summary(g_post_cf)
```

Remember, R-sq is probably so high because size this year is a very good predictor of size next year.

@@

```{r}
g_post_1ha <-
  bam(log_size ~ 
        s(log_size_prev) +
        # s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog_post_1ha,
      select = TRUE,
      method = "REML",
      cluster = cl)
```

@@

```{r}
summary(g_post_1ha)
```

```{r}
draw(g_post_1ha)
```

@@

```{r}
sr <- range(demog_post_cf$log_size_prev)

newdata_covar_cf <- 
  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
         plot = "newplot",
         year_fac = "newyear", 
         spei_history = mean(demog_post_cf$spei_history),
         L = demog_post_cf$L[1:100, ])

@@

covar_growth_cf <-
  ggplot(out, aes(x = exp(log_size_prev), y = exp(.fitted))) + 
  geom_point(data = demog_post_cf, aes(x = size_prev, y = size), alpha = 0.1) +
  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_abline(slope = 1, linetype = 2, color = "red") +
  geom_line(color = "blue") +
  # geom_rug(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.25) +
  theme_linedraw() +
  labs(x = TeX("size_t_"), y = TeX("size_{t+1}_"))
covar_growth_cf
ggsave(here("analysis", "figures", "growth_cf_size.png"), covar_growth_cf)
```
```{r}
out %>%
  filter(exp(log_size_prev) > exp(.fitted)) %>% # shinking plants
  pull(log_size_prev) %>% min() %>% exp() #first size at which shrinking happens.
```

```{r}
sr <- range(demog_post_1ha$log_size_prev)
newdata_covar_1ha <- 
  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
         plot = "newplot",
         year_fac = "newyear",
         spei_history = mean(demog_post_1ha$spei_history),
         L = demog_post_cf$L[1:100, ])

@@

covar_growth_1ha <-
  ggplot(out, aes(x = exp(log_size_prev), y = exp(.fitted))) + 
  geom_point(data = demog_post_1ha, aes(x = size_prev, y = size), alpha = 0.1) +
  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_abline(slope = 1, linetype = 2, color = "red") +
  geom_line(color = "blue") +
  # geom_rug(data = demog_post_1ha, aes(x = size, y = size_next), alpha = 0.25) +
  theme_linedraw() +
  labs(x = TeX("size_t_"), y = TeX("size_{t+1}_"))

covar_growth_1ha
ggsave(here("analysis", "figures", "growth_1ha_size.png"), covar_growth_1ha)
```
```{r}
out %>%
  filter(exp(log_size_prev) > exp(.fitted)) %>% #shrinking plants
  pull(log_size_prev) %>% min() %>% exp() #smallest size where plants start shrinking
```

@@

```{r}
# use mean size plant across both habitats so plots are comparable
ref_data <-
  demog_post %>%
  dplyr::summarize(log_size_prev = log(mean(size_prev, na.rm = TRUE)),  #log of mean size, not mean of log_size
                   year_fac = "newyear",
                   plot = "newlevel")

mean_size <- exp(ref_data$log_size_prev)
```

@@

```{r}
s_post_cf <-
  bam(surv ~ 
        s(log_size_prev) +
        # s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

```{r}
# gam.check(s_post_cf) #not sure what to expect from any of these plots
appraise(s_post_cf)
```

```{r}
summary(s_post_cf)
```

@@

```{r}
s_post_1ha <-
  bam(surv ~ 
        s(log_size_prev) +
        # s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

```{r}
summary(s_post_1ha)
```

@@

covar_s_cf <-
  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_cf) +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  labs(x = TeX("size_t_"), y = "P(survival)") +
  theme_linedraw()

@@

covar_s_1ha <-
  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_1ha) +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  labs(x = TeX("size_t_"), y = "P(survival)") +
  theme_linedraw()

@@

Around 5 percent of plants flowered.

```{r}
f_post_cf <-
  bam(flwr ~ 
        s(log_size_prev) +
        # s(year_fac, bs = "re") + 
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

@@

```{r}
appraise(f_post_cf)
# concurvity(f_post_cf, full = FALSE)
```

these residuals are troubling

@@

This model changes **drastically** depending on whether year is included as a random effect.  Does that mean ther random effect is important? Or is it because it is collinear / concurve with SPEI history?

```{r}
f_post_1ha <-
  bam(flwr ~ 
        s(log_size_prev) +
        # s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

@@

@@

covar_f_cf <-
  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
  stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = demog_post_cf) +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  labs(x = TeX("size_t_"), y = "P(flowering)") +
  theme_linedraw()

@@

covar_f_1ha <-
  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
  stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = demog_post_1ha) +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  labs(x = TeX("size_t_"), y = "P(flowering)") +
  theme_linedraw()

covar_f_1ha
ggsave(here("analysis", "figures", "flwr_1ha_size.png"), covar_f_1ha)
```
